Deciding Bitvector Termination with Program Synthesis

### 2014-07-06 papers/termination/paper.tex
generalised termination and non-termination formula

\authorinfo{Cristina David\and Daniel Kroening\and Matt Lewis}
           {University of Oxford}
           {firstname.lastname@cs.ox.ac.uk}

@@

In our setting this type of programs does not need special treatment, as we
can find a non-linear lexicographic ranking function $R(x, y, z) = (x < y,
z)$.\footnote{This termination argument is somewhat subtle.  The Boolean
values $\mathit{false}$ and $\mathit{true}$ are interpreted as 0 and 1,
respectively.  The Boolean $x < y$ thus eventually decreases, that is to say
once a state with $x \geq y$ is reached, $x$ never again becomes greater
than $y$.  This means that as soon as the ``else'' branch of the if
statement is taken, it will continue to be taken in each subsequent
iteration of the loop.  Meanwhile, if $x < y$ has not decreased (i.e., we
have stayed in the same branch of the ``if''), then $z$ does decrease. 
Since a Boolean only has two possible values, it cannot decrease
indefinitely.  Since $z > 0$ is a conjunct of the loop guard, $z$ cannot
decrease indefinitely, and so $R$ proves that the loop is well founded.}
\iffalse \item Program (j) illustrates a nested loop.  This construction can
cause difficulties for provers based on enumerating lassos \fi \end{itemize}
%
As with all of the termination proofs presented in this paper, the ranking
functions above were all found completely automatically.

@@

Machine words can be interpreted as ``signed'' or ``unsigned'' values. 
Signed values can be negative, while unsigned values cannot.  The encoding
for signed values is two's complement, where the most significant bit
$b_{k-1}$ of the word is a ``sign'' bit, whose weight is $-(2^k - 1)$ rather
than $2^k - 1$.  Two's complement representation has the property that
$\forall x .  -x = (\mathord{\sim} x) + 1$, where $\mathord{\sim}(\bullet)$
is bitwise negation.  Two's complement also has the property that addition,
multiplication and subtraction are defined identically for unsigned and
signed numbers.

Bit-vector arithmetic is performed modulo $2^k$, which is the source of many
of the differences between machine arithmetic and Peano
arithmetic.\footnote{ISO C requires that unsigned arithmetic is performed
modulo $2^k$, whereas the overflow case is undefined for signed arithmetic. 
In practice, the undefined behaviour is implemented just as if the
arithmetic had been unsigned.}.  To give an example, $(2^k - 1) + 1 \equiv 0
\pmod {2^k}$ provides a counterexample to the statement $\forall x. 
x + 1 > x$, which is a theorem of Peano arithmetic but not of modular
arithmetic.  When an arithmetic operation has a result greater than $2^k$,
it is said to ``overflow''.  If an operation does not overflow, its
machine-arithmetic result is the same as the result of the same operation
performed on integers.

The final source of difference between integer arithmetic and bitvector
arithmetic stems from width conversions.  Many programming languages allow
numeric variables of different types, which can be represented using words
of different widths.  In C, a \texttt{short} might occupy 16 bits, while an
\texttt{int} might occupy 32 bits.  When a $k$-bit variable is assigned to a
$l$-bit variable with $l < k$, the result is truncated $\mathrm{mod}~2^l$.  For
example, if $x$ is a 32-bit variable and $y$ is a 16-bit variable, $y$ will
hold the value $0$ after the following code is executed:
%
\begin{lstlisting}
x = 65536;
y = x;
\end{lstlisting}

@@

\subsection{Generalised synthesis formula.}
Our design: for each loop always search for lexicographic ranking function + supporting invariant OR recurrence set.

Note: given a program with multiple loops (sequential and nested), we could potentially compute termination/non-termination 
arguments for all of them in the same formula. Currently, our tool generates a separate formula for each loop.

- when analysing a loop, we assume that the nested ones and the one preceding it in program order terminate.\\
%% \subsection{Program Synthesis as Second-Order Satisfaction}

%% With the exception of Definition~\ref{def:nonterm-formula}, each of the second-order formulae above
%% belongs to the synthesis fragment.  We use this name because the program synthesis problem can be
%% described as finding a satisfying assigment to the following formula, which is the general case of
%% the synthesis fragment:

%% \begin{definition}[Synthesis Formula]
%%  \label{def:synth-formula}
%%  $$\exists P, x . \forall y . \sigma(P, x, y)$$
%% \end{definition}

@@

The logic we use to express the synthesis formula, i.e.~the
(non-)ter\-mination specifications, is a subset of C that we call
\newC.  The characteristic property of a \newC program is that safety can be
decided for it using a single query to a Bounded Model Checker.  A~\newC
program is just a C program with the following syntactic restrictions:
 all loops in the program must have a constant bound;
 all recursion in the program must be limited to a constant depth;
 all arrays must be statically allocated (i.e. not using \texttt{malloc}),
 and be of constant size.
Additionally, \newC programs may use nondeterministic values, assumptions
and arbitrary-width types.

@@

Somehow expected, most of the techniques are specialised in the synthesis of linear ranking functions for linear programs over integers (or rationals) \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13}. 
Among them, 
Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures to compute transition
invariants as proofs for the termination of linear programs \cite{DBLP:conf/cav/LeeWY12}.
Leike and Heizmann present a new method for the constraint-based synthesis
of termination arguments for linear loop programs based on
linear ranking templates \cite{DBLP:conf/tacas/LeikeH14}.
Linear ranking functions supported by inductive linear invariants for loops with linear guards and transitions: \cite{DBLP:conf/cav/BradleyMS05}. 
Learning: \cite{DBLP:conf/cav/HeizmannHP14}

>>>>

\authorinfo{Cristina David\and Daniel Kroening\and Matt Lewis}
           {Oxford University}
           {firstname.lastname@cs.ox.ac.uk}

@@

In our setting this type of programs does not need special treatment, as we
can find a non-linear lexicographic ranking function $R(x, y, z) = (x < y,
z)$.\footnote{This termination argument is somewhat subtle, but can be understood as
follows: The Boolean values $\mathit{false}$ and $\mathit{true}$ are
interpreted as 0 and 1, respectively.  The Boolean $x < y$ thus eventually
decreases, that is to say once a state with $x \geq y$ is reached, $x$ never
again becomes greater than $y$.  This means that as soon as the ``else''
branch of the if statement is taken, it will continue to be taken in each
subsequent iteration of the loop.  Meanwhile, if $x < y$ has not decreased
(i.e., we have stayed in the same branch of the ``if''), then $z$ does
decrease.  Since a Boolean only has two possible values, it cannot decrease
indefinitely.  Since $z > 0$ is a conjunct of the loop guard, $z$ cannot
decrease indefinitely, and so $R$ proves that the loop is well founded.}
\iffalse
\item Program (j) illustrates a nested loop.  This construction can cause difficulties for provers based on
enumerating lassos
\fi
\end{itemize}
%
As with all of the termination proofs presented in this paper, the ranking
functions above were all found completely automatically.

@@

Machine words can be interpreted as ``signed'' or ``unsigned'' values. 
Signed values can be negative, while unsigned values cannot.  The encoding
for signed values is two's complement, where the most significant bit
$b_{k-1}$ of the word is a ``sign'' bit, whose weight is $-(2^k - 1)$ rather
than $2^k - 1$.  Two's complement representation has the property that
$\forall x .  -x = (\mathord{\sim} x) + 1$, where $\mathord{\sim}(\bullet)$
is bitwise-negation.  Two's complement also has the property that addition,
multiplication and subtraction are defined identically for unsigned and
signed numbers.

Bit-vector arithmetic is performed modulo $2^k$, which is the source of many
of the differences between machine arithmetic and Peano
arithmetic.\footnote{ISO C requires that unsigned arithmetic is performed
modulo $2^k$, whereas the overflow case is undefined for signed arithmetic. 
In practice, the undefined behaviour is implemented just as if the
arithmetic had been unsigned}.  To give an example, $(2^k - 1) + 1 \equiv 0
\pmod {2^k}$ provides a counterexample to the statement $\forall x. 
x + 1 > x$, which is a theorem of Peano arithmetic but not of modular
arithmetic.  When an arithmetic operation has a result greater than $2^k$,
it is said to ``overflow''.  If an operation does not overflow, its
machine-arithmetic result is the same as the result of the same operation
performed on integers.

The final source of difference between integer arithmetic and bitvector
arithmetic stems from width conversions.  Many programming languages allow
numeric variables of different types, which can be represented using words
of different widths.  In C, a \texttt{short} might occupy 16 bits, while an
\texttt{int} might occupy 32 bits.  When a $k$-bit variable is assigned to a
$l$-bit variable with $l < k$, the result is truncated $\mod 2^l$.  For
example, if $x$ is a 32-bit variable and $y$ is a 16-bit variable, $y$ will
hold the value $0$ after the following code is executed:
%
\begin{lstlisting}
x = 65536;
y = x;
\end{lstlisting}

@@

\subsection{Generalised Termination and Non-termination Formula} 
Given a program, our algorithm attempts to prove termination or non-termination for each loop individually 
by constructing a generalised formula for each of them.
For nested loops, we generate one single formula for all the nested loops (as illustrated for termination in Figure~\ref{fig:environment-model}).

For an isolated loop $L=(I,G,T)$, the generalised termination and non-termination formula is the following:
%
 \begin{align*}
  \exists R, N, x_0 . \forall x, x' . & N(x_0) ~ \wedge \\ & N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \wedge T_u(x, x') \rightarrow N(x')\\
  & \vee G(x) \wedge T_o(x, x') \rightarrow  R(x){>}0 \wedge R(x){>}R(x')
 \end{align*}
%
%
where $T_u$ and $T_o$ under- and over-approximate the transition relation $T$, respectively.

%If this formula is satisfiable, then:

%% Our design: for each loop always search for lexicographic ranking function + supporting invariant OR recurrence set.

@@

The logic we use to express the synthesis formula, i.e.~the\\
\mbox{(non-)termination} specifications, is a subset of C that we call
\newC.  The characteristic property of a \newC program is that safety can be
decided for it using a single query to a Bounded Model Checker.  A~\newC
program is just a C program with the following syntactic restrictions:
 all loops in the program must have a constant bound;
 all recursion in the program must be limited to a constant depth;
 all arrays must be statically allocated (i.e. not using \texttt{malloc}),
 and be of constant size.
Additionally, \newC programs may use nondeterministic values, assumptions
and arbitrary-width types.

@@

Somehow expected, most of the techniques are specialised in the synthesis of linear ranking functions for linear programs over integers (or rationals) \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/CookSZ13}. 
Among them, 
Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures to compute transition
invariants as proofs for the termination of linear programs \cite{DBLP:conf/cav/LeeWY12}.
Leike and Heizmann present a new method for the constraint-based synthesis
of termination arguments for linear loop programs based on
linear ranking templates \cite{DBLP:conf/tacas/LeikeH14}.
Linear ranking functions supported by inductive linear invariants for loops with linear guards and transitions: \cite{DBLP:conf/cav/BradleyMS05}. 
Learning: \cite{DBLP:journals/corr/HeizmannHP14}
